---
title: "SLE777_Assignment4"
author: "tdxcn"
date: "`r Sys.Date()`"
output: html_document
---
1. Download raw data of "gene_expression" file 
```{r}
download.file("https://raw.githubusercontent.com/ghazkha/Assessment4/main/gene_expression.tsv",destfile="gene_expression.tsv") #download the "gene_expression" file from Mr.Ghazkha github
```
```{r}
gene_expression <- read.delim("gene_expression.tsv",header=TRUE) #import the downloaded "gene_expression" file into the local Rstudio in form of table 
str(head(gene_expression)) #showing the structure of data in "gene_expression"
```
Making the gene identifier the row name 
```{r}
rownames(gene_expression) <- gene_expression$Name_Description #replacing the row names in "gene_expression" with GeneIDs
gene_expression$Name_Description <- NULL #Delete the existing GeneID column
```

2. Creating row means column and showing a table of values for the first six genes
```{r}
gene_expression$Mean_expression <- rowMeans(gene_expression) #Creating a new column named "Mean_expression" and assigning it with mean value of columns in each row 
gene_expression$Mean_expression <- round(gene_expression$Mean_expression, digits=2) #Rounding up the value in "mean_expression" column to two decimal numbers
head(gene_expression,6) # Printing the first 6 genes
```
3. Listing the 10 genes with the highest mean expression
```{r}
gene_expression <- gene_expression[order(gene_expression$Mean_expression,decreasing = TRUE),] #Reordering the rows in "gene_expression" by the order of the values in "Mean_expression" from largest to smallest 
head(gene_expression,10,drop=FALSE) #Printing a table of the first 10 genes with the highest mean expression
```
4. The number of genes with a mean expression lower than 10
```{r}
sum(gene_expression$Mean_expression < 10) #Counting the number of gene with value in "Mean_expression" lower than 10
```
5. Making a histogram of mean values
```{r}
hist(gene_expression$Mean_expression,#Making histogram plot using values in the "Mean_expression" column
     breaks = 1000, #Adjusting the number of bin in the histogram to 1000  
     col = "blue",#Setting to color of columns to blue
     border = "black",#Setting the color of the borders to black
     main = "Histogram of Mean Expression Values", #Setting the name of the histogram 
     xlab = "Mean Expression", #Setting the name of x-axis 
     ylab = "Frequency", #Setting the name of y-axis
     ylim = c(0, 60000), #Setting the limit of y-axis to 60000
     xlim = c(0,60000)) #Setting the limit of y-axis to 60000
```
From the histogram, it can be observed that a majority of genes (around 46000) showed mean gene expression within the range from 0 to 10. Above the range from 0 to 10, the frequency of mean expression drasstically decrease as their value increase. 
 


