---
title: "Part1_growth_data"
author: "tdxcn"
date: "2024-10-09"
output: html_document
---
6. Download and import the "growth_data" file
```{r}
download.file("https://raw.githubusercontent.com/ghazkha/Assessment4/main/growth_data.csv",destfile="growth_data.csv") #download the "growth_data" file from Mr.Ghazkha github
growth_data <- read.csv("growth_data.csv",sep=",", header = TRUE) #import the downloaded "growth_data" file into the local Rstudio as a csv file
str(head(growth_data)) #showing the structure of data in "growth_data"
```

Showing column name of "growth_data" file
```{r}
colnames(growth_data) #Presenting the column names
```
7. Calculating row means and standard deviation
```{r}
growth_data$Mean_growth <- rowMeans(growth_data[,3:6]) #Creating a "Mean_growth" column in "growth_data" and assigning it with mean values of columns in each row  
growth_data$Standard_deviation <- apply(growth_data[,3:6], 1, sd, na.rm = TRUE) #Creating a "Standard_deviation" column in "growth_data" and assigning it with standard deviation values of columns in each row  
growth_data$Standard_deviation <- round(growth_data$Standard_deviation, digits=2) #Rounding up the values in "Standard deviation" to 2 decimals 
```

Grouping data by Site
```{r}
growth_data <- subset(growth_data[grouping(growth_data$Site == "northeast"),]) #Grouping the values rows in "growth_data" into two sections, one has "southwest" Site and the other has "northeast" Site
```

8. Box plot of tree circumference at the start and end of the study at both sites
```{r}
boxplot(growth_data[,3:6], xlab = "Year", ylab = "Growth", main="Tree circumference at the start and end of the study", ylmit = 200) #Making a box plot of the circumference at the beginning and the end of the study at both site
```
9. Mean growth over the last 10 years at each site 
Creating subsets for data of trees grown in southwest and northeast
```{r}
southwest_data <- growth_data[growth_data$Site == "southwest",] #Making a subset of data of trees in the southwest site
northeast_data <- growth_data[growth_data$Site == "northeast",] # Making a subset of data of trees in the northeast site
southwest_data$Mean_growth <- NULL #deleting the "Mean_growth" column
northeast_data$Mean_growth <- NULL #deleting the "Mean_growth" column
southwest_data$Standard_deviation <- NULL #deleting the "Standard_deviation" column
northeast_data$Standard_deviation <- NULL #deleting the "Standard_deviation" column
rownames(southwest_data) <- c(1:nrow(southwest_data)) #Renumber the the row in "southwest_data" following the order from smallest to largest
rownames(northeast_data) <- c(1:nrow(northeast_data)) #Renumber the the row in "northeast_data" following the order from smallest to largest
southwest_data$Mean_growth_last_10_years <- rowMeans(southwest_data[,4:6]) #Calculating mean growth of trees in southwest site over the last years of the study 
northeast_data$Mean_growth_last_10_years <- rowMeans(northeast_data[,4:6]) #Calculating mean growth of trees in northeast site over the last years of the study
southwest_data$Mean_growth_last_10_years <- round(southwest_data$Mean_growth_last_10_years, digits = 2) #rounding up the mean values to 2 decimals 
northeast_data$Mean_growth_last_10_years <- round(northeast_data$Mean_growth_last_10_years, digits = 2) #rounding up the mean values to 2 decimals 
```

10. t test of the mean growth value over the last 10 year at the two sites
H0: there is no significant difference between the mean growth over the last 10 years of the study at each site
H1: there is a significant difference between the mean growth over the last 10 years of the study at each site 
```{r}
t_test_result <- t.test(southwest_data$Mean_growth_last_10_years,northeast_data$Mean_growth_last_10_years) #Performing t test on the mean growth value of two sites
```
The t.test results in a p-value of 0.042 , which is smaller than the 0.05 level of significance.
Therefore, the hypothesis H1 is accepted
Thus, there is a significant difference in the mean growth of trees at each site over the last 10 years of the study 