---
title: "Part1_growth_data"
author: "tdxcn"
date: "2024-10-09"
output: html_document
---
Download the file 
```{r}
download.file("https://raw.githubusercontent.com/ghazkha/Assessment4/main/growth_data.csv",destfile="growth_data.csv")
```

Column name in growth_data
```{r}
colnames(growth_data)
```
Calculating row means and standard deviation
```{r}
growth_data$Mean_growth <- rowMeans(growth_data[,3:6])
growth_data$Standard_deviation <- apply(growth_data[,3:6], 1, sd, na.rm = TRUE)
growth_data$Standard_deviation <- round(growth_data$Standard_deviation, digits=2)
```

Grouping data by Site
```{r}
growth_data <- subset(growth_data[grouping(growth_data$Site == "northeast"),]) 
```

Box plot of tree circumference at the start and end of the study at both sites
```{r}
boxplot(growth_data[,3:6], xlab = "Year", ylab = "Growth", main="Tree circumference at the start and end of the study at both sites")
```
Creating subsets for data of trees grown in southwest and northeast
```{r}
southwest_data <- growth_data[growth_data$Site == "southwest",]
northeast_data <- growth_data[growth_data$Site == "northeast",]
```

Mean growth of each tree at each site has already been integrated into the table 

t.test to estimate the p-value that the 10 year growth is different at the two sites
H0: there is no significant difference between the mean growth over the years at each site 
H1: there is a significant difference between the mean growth over the years at each site 
```{r}
t_test_result <- t.test(southwest_data$Mean_growth,northeast_data$Mean_growth)
```
The t.test results in a p-value of 0.0389 , which is substantially smaller than the 0.05 level of significance.
Therefore, the hypothesis H1 is accepted
Thus, there is a significant difference in the mean growth of trees at each site over the years 